Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-15 appendix/appendix.tex
New para
New fix

Local development and some runs of small size were done under the hardware listed in Table \ref{table:local_hardware}. When running ABC SMC, the program can make used of 8 python process (Hyper-Threading enabled) and run under 3.8GHz (maximal) for a long time under proper cooling, proved that the program could efficiently use the computation resources for a local personal computer. It is noted that the program took no advantage of graphics card, as \verb|multiprocessing| along can out exploit the resources inside CPU. The execution time and other performance data presented in Chapter 5 were obtained using remote machines but not local machine.

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter            & Value & Unit                     \\ [0.5ex]
        \hline\hline
        $\lambda_N$          & 2.20  & $cell/h$                 \\
        $\kappa_{N\beta}$    & 3.96  & $cell/(unit\cdotp h)$    \\
        $\mu_N$              & 1.72  & $h^{-1}$                 \\
        $\nu_{N\Phi}$        & 0.220 & $cell^{-1}\cdotp h^{-1}$ \\
        \hline
        $\lambda_\Phi$       & 1.31  & $cell/h$                 \\
        $\kappa_{\Phi\beta}$ & 0.124 & $cell/(unit\cdotp h)$    \\
        $\mu_\Phi$           & 0.145 & $h^{-1}$                 \\
        \hline
        $s_{\beta N}$        & 6.554 & $unit/(cell\cdotp h)$    \\
        $i_{\beta\Phi}$      & 1.71  & $cell^{-1}$              \\
        $\mu_\beta$          & 0.521 & $h^{-1}$                 \\
        \hline
        $s_{\alpha\Phi}$     & 10.2  & $unit/(cell\cdotp h)$    \\
        $\mu_\alpha$         & 19.1  & $h^{-1}$                 \\
        \hline
    \end{tabular}
    \caption{Parameter values used for model 1}
    \label{table:synthetic_data}
\end{table}

@@

>>>>

Local development and some runs of small size were done under the hardware listed in Table \ref{table:local_hardware}. When running ABC SMC, the program can make used of 8 python process (Hyper-Threading enabled) and run under 3.8GHz (maximal) for a long time under proper cooling, proved that the program could efficiently use the computation resources for a local personal computer. It is noted that the program took no advantage of graphics card, as \verb|multiprocessing| along can only exploit the resources inside CPU. The execution time and other performance data presented in Chapter 5 were obtained using remote machines but not local machine.

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|c c|}
        \hline
        Parameter            & Value \\[0.5ex]
        \hline\hline
        $\lambda_N$          & 2.20  \\
        $\kappa_{N\beta}$    & 3.96  \\
        $\mu_N$              & 1.72  \\
        $\nu_{N\Phi}$        & 0.219 \\
        \hline
        $\lambda_\Phi$       & 1.31 \\
        $\kappa_{\Phi\beta}$ & 0.124 \\
        $\mu_\Phi$           & 0.145 \\
        \hline
        $s_{\beta N}$        & 6.55  \\
        $i_{\beta\Phi}$ & 1.71 \\
        $\mu_\beta$          & 0.521  \\
        \hline
        $s_{\alpha\Phi}$     & 10.2 \\
        $\mu_\alpha$         & 19.7  \\
        \hline
    \end{tabular}
    \caption[TODO]
    {Known values TODO}
    \label{table:known_values}
\end{table}

@@

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/inferno5.png}}
    \end{center}
    \caption[Joint density distribution of each parameter pair]%
    {Joint distribution of each parameter pair. MORE DESCRIPTION}
    \label{fig:model5_para_joint}
\end{figure}

### 2020-08-15 implementation/implementation.tex
New fix

Estimation of  parameters in the proposed models is the main task of this section, which involves the implementation of ABC SMC and several experiments. ABC SMC-based model selection is another task that can be easily conducted with the developed parameter inference framework.

The build and test of code were under \verb|Python| environment with \verb|pyABC|\cite{ref:pyabc}. Some other code e.g. shell script and \verb|R| code are also used to perform the experiments and analyse the results. Software and system environment used in our implementation can be found at Appendix A. The build and test were performed on local computers and HPC facilities available within EPCC.

\section{Implementations and code}

 [ABC implementations details, e.g. distance, population, ODE related functions]

 [how the code is developed and built]

According to our preliminary studies, ABC SMC framework implemented using \verb|pyABC| in \verb|Python| was adopted in this project. \verb|pyABC| is a popular SMC software packages \cite{ref:pyabc} that has been used form many SMC studies and applications \cite{population}. \verb|pyABC| provides an open-source framework for likelihood-free inference, which is an implementation of ABC SMC algorithm and a tool-box for our own inference and analysis tasks. Besides, it supports multi-core execution implemented using \verb|multiprocessing| package for scaling-up, which is of our interests and suitable for the followed performance experiments.

The inference framework code for the project was developed and tested in local environment (macOS 10.15.6, see Table \ref{table:local_macine}) at first, then the functional version was deployed on compute node of ARCHER and Cirrus for parameter inference experiments and model selection task. The scaling-up experiment was performed on Cirrus, and the profiling and its analysis were performed on local machine. Some development and debug were on the remote machines to ensure experiments run correctly on compute node.

The ODE model-related code and utilities were firstly developed to enable reusable functions, variables and data calls. It includes the code for the ODE models, ODE solver, data structures, data format transform functions etc. By doing this in a separate code file, we can change options and activate ABC SMC easily in the implementation code file, without creating definition and duplicated code for the models and parameter priors.

Some modifications were made to the source code of \verb|pyABC|. For example, the laboratorial measured data were not considered `complete' for modeling and analysis purpose: the available time points for cells and cytokines are different, e.g. cell number was not measured at 0.25 hpl and cytokines' expression was not measured at time point 120 hpl. These missing values were treated as `ignored' when calculating the distance between observed data and simulated data. To cope with this built-in distance function code is modified.

Analysis code was also wrapped up into functions which can be easily called after an ABC SMC run is finished, which includes multiple visualisations and summary statistic of the results. For different tasks there are also separated analysis code file. Other analysis tools used in the project include built-in database visualisation tool, R code and Microsoft Excel.

@@

\section{ABC SMC and model hyperparameters experiments}

As the key focus of this project is on the parameter estimations of dynamical system, the ability of inference on the proposed models was firstly examined before actual inference using the experimental observed data.

@@

    [including the options that are tested/ tried/ not adopted]

    [including experiments results and discussions]

    [how do these options correspond to biological model and how to set them accordingly for proposed model]

The ABC SMC inference can have largely different performance when using different hyperparameters and implementation options. To firstly test the ability of inference and observe the results of different options, synthetic data with known true parameter values are used as target data. The true value of parameters that are used to generate synthetic data is listed in Appendix B.1. These true values are firstly obtained by fitting a least square fit of model 1 onto the experimental data (Figure \ref{fig:infer_back_data}), the target data is then generated using these parameter values via ODE solver in \verb|Python|.

\begin{figure}[h!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/LS.png}}
    \end{center}

@@

[TABLE for known parameter values]

[why use synthetic data]

@@

[how adaptive distance work]

@@

@@

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/ib.png}}
    \end{center}

@@

Two types of factors and adaptive distance function are tested by running the same number of generations (20 generation, 2000 particles per generation). From FIGURE, by applying factors the total required samples are less, and the adaptive distance requires much less samples; their acceptance rates are also higher than the standard one. It seems that adaptive function and factors could help to converge more quickly.

However considering the result after 20 generation (FIGURE), the resultant model gives the contrary preference. Applying factors and using adaptive distance can lead the resultant model less accurate (after the same generation iterations): they have wider inter-quantile range and may requires more generations to converge. TODO

\subsubsection{Conclusion} We noticed the significant efficiency improvement of these adaptive options; they can largely improve the acceptance rate in some cases. However, as the factors and weights are directly multiplied to $\Delta x$ and thus the metrics for distance are changed, a direct compare of the efficiency under the same threshold schedule is unfeasible; after the same number of generations they can reach a small $\epsilon_t$ but still the approximated posterior is not as accurate as the standard implementation. AS a result, our implementation of the parameter inference will only consider standard distance function and factors are only tested for improvements (FIGURE).

@@

Regarding the dynamic systems model, SMC was applied with two data size options, three prior distribution (uniform distribution, wider uniform distribution and log-uniform distribution). The population options i.e. population size and number of populations were also tried with different values. These experiments intended to give suggestions on the later SMC implementation on experimental data.

@@

    [discuss uniform and log uniform when result ready]

\begin{figure}
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/size1.png}}
    \end{center}

@@

[FIGURE HERE]

@@

    \caption[Model comparison of model 1, 2 and 3]%
    {Model comparison of model 1, 2 and 3. From population 8 to population 30, model3 always wins}
    \label{fig:cmp1}

@@

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/inferno5.png}}
    \end{center}
    \caption[Joint density distribution of each parameter pair]%
    {Joint distribution of each parameter pair. MORE DESCRIPTION}
    \label{fig:model5_para_joint}
\end{figure}

@@

@@

\begin{figure}[t]
    \begin{center}
        \resizebox{0.8\hsize}{!}{\includegraphics{fig/pc_pie_all.png}}
    \end{center}

@@

\begin{figure}[h]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/pc.png}}
    \end{center}

    \caption[PCA results of the last population of model 5]{PCA results of the last population of model 5. MORE DESCRIPTION}
    \label{fig:pc}
\end{figure}

>>>>

For the proposed models, exploration of the uncertainties in the parameter estimations is our main task in this chapter, which involved the application of ABC SMC and several experiments. After that, the parameter inference framework was also used for model selection.

The build and test of the code were under \verb|Python| environment with \verb|pyABC|\cite{ref:pyabc}. Some other code, e.g. shell scripts and \verb|R| code were also used to perform the experiments and result analysis. Software and system environment used in our implementation can be found in Appendix A. The build and test were performed on local computers and HPC facilities available within EPCC.

\section{Code and implementations}

%  [ABC implementations details, e.g. distance, population, ODE related functions]

%  [how the code is developed and built]

According to our preliminary studies, a ABC SMC framework implemented using \verb|pyABC| in \verb|Python| was adopted in this project. \verb|pyABC| is a popular SMC software packages\cite{ref:pyabc} which has been used form many ABC SMC inference applications\cite{population}. \verb|pyABC| provides an open-source framework for likelihood-free inference, which is a \verb|NumPy| implementation of ABC SMC algorithm and a useful tool-box for our own inference and analysis tasks. Besides, it supports multi-core execution implemented using \verb|multiprocessing| built-in package, and it made our further studies in the scaling-up performance possible.

%  which is of our interests and suitable for the followed performance experiments.

The inference framework code for the project was developed and tested in local environment (macOS 10.15.6, see Table \ref{table:local_macine}) at first using a small input, then the functional version was deployed on compute node of ARCHER and Cirrus for parameter inference experiments and model selection tasks. The scaling-up experiment was performed on Cirrus. The profiling and its analysis were performed on local machine. Some necessary development and debug were performed on the remote machines to ensure that experiments run correctly on compute node.

The ODE model-related code and utilities were firstly developed to enable reusable functions, variables and data calls. It included the code for the ODE equations, ODE solver, data structures and data format transform functions, etc. By doing this in separate code files, we could change implementation options and activate ABC SMC easily in another `trigger' file, without creating definition and duplicated code for the models, observed data and parameter priors.

Some missing features were found when using \verb|pyABC|, therefore some modifications were made to the source code of \verb|pyABC|.

For example, the laboratorial measured data were not considered `complete' or `tidy' for modeling and analysis purpose: the available time points for cells and cytokines were different, e.g. cell number was not measured at 0.25 hpl and cytokines' expression was not measured at time point 120 hpl. These missing values were treated as `ignored' when calculating the distance between observed data and simulated data, and represented by \verb|NaN|s. However, in some distance calculations \verb|NaN| was not acceptable; to cope with this, built-in distance function was modified. Other modified code include some visualisation and result representation enhancements

Analysis of the program output was also wrapped up into functions which can be easily called after an ABC SMC run is finished, which includes multiple visualisations and summary statistic of the results. For different tasks there are also separated analysis code file. Other analysis tools used in the project include built-in database visualisation tool, R code and Microsoft Excel.

@@

\section{Hyperparameters and implementation options experiments}

As the key focus of this project was on the parameter estimations of dynamical system, the ability of inference on the proposed models was firstly examined before actual inference using the experimental observed data.

@@

    % [including the options that are tested/ tried/ not adopted]

    % [including experiments results and discussions]

    % [how do these options correspond to biological model and how to set them accordingly for proposed model]

    % [why use synthetic data]

The ABC SMC inference can have largely different performance when using different hyperparameters and implementation options. To firstly test the ability of inference and observe the results of different options, synthetic data with known true parameter values are used as target data. by doing this, we could directly compare the inferred parameters to the true value, and compare the simulated trajectories to the true trajectories and then observe the result to see whether the inference is successful and how well the model fit the data. Also, by setting the same final threshold value, the efficiency under different options can be compared and help us to choose proper options when applying ABC SMC on the observed data. The true value of parameters that were used to generate synthetic data is listed in Table \ref{table:known_values}.

The true values were obtained by fitting model 1 onto the experimental data (Figure \ref{fig:infer_back_data}) using least-square optimisation method. Available least-square fit tools in \verb|SciPy| were used. The fit algorithm was Trust Region Reflective\footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.optimize.least_squares.html}}. Target data was then generated using these parameter values via ODE integrate solver\footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.odeint.html}}.

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/LS.png}}
    \end{center}

@@

@@

% [how adaptive distance work]

@@

\subsubsection{Experiment results}

@@

Two types of factor and adaptive distance function were tested by running the same number of generations (20 generation, 2000 particles per generation). The first type of factor is 25:75  factor, where first half of the trajectory data was assigned higher factor (75), and second half of the data was assigned lower factor (25); this could make the first half if the data more `important' in the inference. The second factor was range factor, which tried to normalise the trajectory of each dependent variable according to its data range.

From Figure \ref{fig:factor}, different types of factor gave different performance. 25:75 factor required much more samples in the first generation (orange bar in the left of Figure \ref{fig:factor}) and also did not improve much in later generations, thus it is the slowest one. The first half of observed data contained more features and information, assigning it with higher factor would make the algorithm think that it is more important to fit the first half and thus focus on the hard part, and consequently more particles were needed in the inference process. The acceptance rates of 25:75 factor among different generations were also generally slightly lower than the standard. By contrast, range factor required less total samples and gave generally higher acceptance rates than standard, which made it considerably faster. Adaptive distance gave the best efficiency, as it significantly improved the acceptance rates and resulted in even less required samples (right, Figure \ref{fig:factor}).

\begin{figure}[t!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/ib.png}}
    \end{center}

@@

However, the result of factors adaptive distance could not be directly compared to standard run, because it changed the distance measurement by adapting weights or manually setting factors. Factors were normalise to make it more close the standard distance measurement. As a result, the same final threshold value did not guarantee that adaptive distance can reach the same level of fit as others. After 20 populations, adaptive distance had a poor fit (Figure \ref{fig:factor_sim}), while the standard run and two types of factors had similar tight fits of the data, which all converge to the true posterior. Additional in our implementations, variance factor (which tries to normalise the data according to the dependent variables' variance) was also tries and gave very similar performance as range factor.

To conclude, because the factors and weights are directly multiplied to $\Delta x$ and thus the metrics for distance are changed, the direct compare of the efficiency under the same threshold schedule was unfeasible; we could only compare the results after the same number of generations. Factors gave nearly the same inferred results but required different number of samples, hence cares must betaken when choosing factor schemes. Adaptive distance did not result in a accurate result as others. As a result, our further implementations used standard distance function, and factors were tried for possible improvements.

% by applying factors the total required samples are less, and the adaptive distance requires much less samples; their acceptance rates are also higher than the standard one. It seems that adaptive function and factors could help to converge more quickly.

% However considering the result after 20 generation (Figure \ref{fig:factor_sim}), the resultant model gives the contrary preference. Applying factors and using adaptive distance can lead the resultant model less accurate (after the same generation iterations): they have wider inter-quantile range and may requires more generations to converge. TODO

% \subsubsection{Conclusion} We noticed the significant efficiency improvement of these adaptive options; they can largely improve the acceptance rate in some cases. However, as the factors and weights are directly multiplied to $\Delta x$ and thus the metrics for distance are changed, a direct compare of the efficiency under the same threshold schedule is unfeasible; after the same number of generations they can reach a small $\epsilon_t$ but still the approximated posterior is not as accurate as the standard implementation. AS a result, our implementation of the parameter inference will only consider standard distance function and factors are only tested for improvements (FIGURE).

@@

Regarding the existing dynamic systems model, SMC was applied with two data size options, three prior distribution (uniform distribution, wider uniform distribution and log-uniform distribution). The population options i.e. population size and number of populations were also tried with different values. These experiments intended to give suggestions on the later SMC implementation on experimental data.

@@

\begin{figure}[t!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/size1.png}}
    \end{center}

@@

%     [discuss uniform and log uniform when result ready]

% [FIGURE HERE]

@@

    \caption[ABC SMC based model comparison of model 1, 2 and 3]%
    {ABC SMC based model comparison of model 1, 2 and 3. From population 8 to population 30, model3 always wins}
    \label{fig:cmp1}

@@

@@

\begin{figure}[H]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/model345cmp.png}}
    \end{center}

    \caption[ABC SMC based model comparison of model 3, 4 and 5]{ABC SMC based model comparison of model 3, 4 and 5. MORE DETAIL}
    \label{fig:overfit}
\end{figure}

@@

\begin{figure}[H]
    \begin{center}
        \resizebox{0.8\hsize}{!}{\includegraphics{fig/pc_pie_all.png}}
    \end{center}

@@

\begin{figure}[H]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/pc.png}}
    \end{center}

    \caption[PCA results of the last population of model 5]{PCA results of the last population of model 5. MORE DESCRIPTION}
    \label{fig:pc_pie}
\end{figure}
